---
title: "Reproducible_Research_Assign1"
author: "Kobra"
date: "17/06/2019"
output: 
  html_document: 
    keep_md: yes
---
# Reproducible Research Course
## Assignment 1

Installing required packages if needed:
``` {r echo= FALSE}
library(dplyr)
#install.packages("lattice")
library(lattice)
```

Loading data into R:

```{r loading the data}
URL <-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file (URL, destfile= "./zip_file.zip", mode="wb")
unzip ("zip_file.zip")
mydata <- read.csv ("activity.csv")
summary (mydata)
```


Total of steps taken per day calculation and its histogram:

``` {r}
mydata$date <- as.Date(mydata$date)

total_steps_per_day <- mydata %>% group_by(date) %>% summarise(sum_steps=sum(steps, na.rm = TRUE))

head (total_steps_per_day)

png("Fig1.png")
hist (total_steps_per_day$sum_steps, xlab="Total steps taken per day", ylab="", main="Frequency of total steps per day")
dev.off()

hist (total_steps_per_day$sum_steps, xlab="Total steps taken per day", ylab="", main="Frequency of total steps per day")

```

Median and mean values for total steps taken per day:
``` {r mean and median of total steps per day}
total_steps_per_day %>% summarise (mean_tspd =mean (sum_steps, na.rm=TRUE), median_tspd = median (sum_steps, na.rm=TRUE))
```

Time series plot for total steps per interval:
```{r}

five_min_interval <- format( seq.POSIXt(as.POSIXct(Sys.Date()), as.POSIXct(Sys.Date()+1), by = "5 min"), "%H%M", tz="GMT")
five_min_interval <- five_min_interval [-1]
five_min_interval [288] <- "2400"


steps_groupedby_interval <- mydata %>% group_by(date) %>% mutate (interval=five_min_interval)
steps_groupedby_interval_averaged <- steps_groupedby_interval %>% group_by(interval) %>% summarise(average_intervalled=mean (steps, na.rm=TRUE))
head (steps_groupedby_interval_averaged)

png("Fig2.png")

plot (x=steps_groupedby_interval_averaged$interval, y=steps_groupedby_interval_averaged$average_intervalled, type="l", xlab= "5 minute intervals", ylab ="Average steps per interval")

dev.off()

plot (x=steps_groupedby_interval_averaged$interval, y=steps_groupedby_interval_averaged$average_intervalled, type="l", xlab= "5 minute intervals", ylab ="Average steps per interval")
```

The interval in which the maximum average number of steps is recorded.
``` {r}
steps_groupedby_interval_averaged$interval[which.max(steps_groupedby_interval_averaged$average_intervalled)]
```

Missing values:

```{r} 
total_NA <- sum (is.na (mydata [, "steps"]))
total_NA
```

Replacing NAs using the average cacluated fr each interval:

```{r}
na_replaced <- replace (steps_groupedby_interval$steps, is.na (steps_groupedby_interval$steps), steps_groupedby_interval_averaged$average_intervalled)

steps_groupedby_interval$steps_na_replaced <- na_replaced
head (steps_groupedby_interval)
```

Calculating the total number of steps taken per day after replacing the NAs as well as mean adn median:

``` {r}
total_steps_per_day_2 <- steps_groupedby_interval %>% group_by(date) %>% summarise(sum_steps_2=sum(steps_na_replaced))
head (total_steps_per_day_2)

png("Fig3.png")
hist (total_steps_per_day_2$sum_steps_2, xlab="Total steps taken per day", ylab="", main= "Frequency of total steps per day")
dev.off()

hist (total_steps_per_day_2$sum_steps_2, xlab="Total steps taken per day", ylab="", main= "Frequency of total steps per day")

total_steps_per_day_2 %>% summarise (mean_tspd2 =mean (sum_steps_2), median_tspd2 = median (sum_steps_2))

```

The mean and median both seems to be increasing. The reason could be due to the fact that in the first round we were omitting the NAs from the calculation, however, now they have a value which is equal to the average number of steps for each interval. This will skew the result to have a higher mean value.

Comparing the activity pattern on weekdays versus weekends:

```{r}
steps_groupedby_interval$weekday <- weekdays (steps_groupedby_interval$date)

steps_groupedby_interval$weekday.type <- ifelse(steps_groupedby_interval$weekday == "Saturday" | steps_groupedby_interval$weekday == "Sunday", "Weekend", "Weekday")
steps_groupedby_interval$weekday.type <- factor (steps_groupedby_interval$weekday.type)
head (steps_groupedby_interval)

```

```{r}
steps_groubedBy_interval_weekday <- steps_groupedby_interval %>% group_by (interval, weekday.type) %>% summarise(mean_sgiw = mean (steps, na.rm =TRUE))


steps_groubedBy_interval_weekday$interval <- as.numeric(as.character(steps_groubedBy_interval_weekday$interval))

png("Fig4.png")

xyplot (mean_sgiw ~ interval|weekday.type, steps_groubedBy_interval_weekday, type= "l", layout = c(1, 2), xlab = "5 minute intervals", ylab = "Average steps per interval")
dev.off()

xyplot (mean_sgiw ~ interval|weekday.type, steps_groubedBy_interval_weekday, type= "l", layout = c(1, 2), xlab = "5 minute intervals", ylab = "Average steps per interval")

```

